\section{Integrating Neuro-Symbolic GNNs with Inclusion Maps and Vector Clocks for High-Frequency Market Event Modeling}

This approach combines Graph Neural Networks (GNNs), symbolic temporal logic, and measure-theoretic constructs to model and analyze high-frequency market events. It treats market data---such as trades, price updates, and macroeconomic signals---as events embedded in a measurable event space, using \emph{inclusion maps} to structure their relationships and \emph{vector clocks} to capture partial orderings across distributed agents. The system is designed to address the limitations of purely numerical or purely symbolic systems by integrating the strengths of both paradigms.

\subsection{Modeling Framework}

We define a measurable event space $(\Omega, \mathcal{F})$, where:
\begin{itemize}
    \item $\Omega$: The sample space of all possible market events.
    \item $\mathcal{F}$: A $\sigma$-algebra over $\Omega$, containing measurable subsets such as the set of all trades at time $t$, or all events involving a specific security.
\end{itemize}

Each \emph{agent} (e.g., trader, exchange, algorithm) observes a subset of $\mathcal{F}$, denoted $\mathcal{F}_i \subseteq \mathcal{F}$, and each such subset is connected to others via \emph{inclusion maps} $i_{ij}: \mathcal{F}_i \hookrightarrow \mathcal{F}_j$, forming a presheaf over the category of agents. This structure models partial observability and information flow in the market.

\subsection{Temporal Logic with Vector Clocks}

Let each event $e$ carry a \emph{vector clock} $v(e) \in \mathbb{N}^n$, where $n$ is the number of agents. Define the partial ordering $e \rightarrow e'$ (i.e., ``event $e$ happened before $e'$'') as:
\[
v(e) < v(e') \iff \forall i,\ v(e)_i \leq v(e')_i \ \text{and} \ \exists j \ \text{s.t.} \ v(e)_j < v(e')_j
\]
This captures \emph{causality} in the system---crucial for determining arbitrage paths, information flow, or detecting violations of market fairness.

\subsection{Graph Neural Network Integration}

We define a graph $G_t = (V_t, E_t)$ at time $t$, where:
\begin{itemize}
    \item $V_t$: Vertices represent market agents or securities.
    \item $E_t$: Edges represent causal or informational dependencies between entities, computed via inclusion maps and vector clock comparisons.
\end{itemize}

Each node $v \in V_t$ has a feature vector $x_v$, incorporating:
\begin{itemize}
    \item Observed events (trades, quotes, news),
    \item Historical state (price velocity, momentum, etc.),
    \item Vector clock embeddings (relative to other agents).
\end{itemize}

The GNN message-passing function $\phi$ respects the partial ordering induced by the vector clocks:
\[
h_v^{(k+1)} = \text{UPDATE}^{(k)}\left(h_v^{(k)}, \sum_{\substack{u \in \mathcal{N}_v^{(k)} \\ v(u) < v(v)}} \text{MESSAGE}^{(k)}(h_u^{(k)}, h_v^{(k)}, e_{uv})\right)
\]
This ensures that information from temporally prior (but possibly concurrent) events is correctly propagated, enabling \emph{non-linear temporal reasoning} across the network.

\subsection{Symbolic-Probabilistic Fusion}

Each GNN layer outputs a distribution over measurable subsets of $\mathcal{F}$, representing updated \emph{beliefs} about latent variables like volatility spikes, manipulation risk, or systemic shocks. These subsets can be mapped back to symbolic expressions in first-order temporal logic, e.g.:

\begin{quote}
``There exists a trade sequence from agent A to B such that the price impact exceeds threshold $\delta$ within $\tau$ ms.''
\end{quote}

Thus, predictions are not just pointwise but \emph{measurable set-valued}, and interpretable.

\subsection{Applications}

\begin{itemize}
    \item \textbf{Arbitrage Detection}: Detect cycles in the event graph that violate causality or information asymmetry constraints.
    \item \textbf{Latency Arbitrage Immunization}: Identify vulnerable edges (slow links) and symbolically verify resilience via inclusion-preserving rewrites.
    \item \textbf{Market Manipulation Detection}: Capture spoofing or layering patterns as fixed points or anomalies in vector-clock-ordered subgraphs.
    \item \textbf{Macro-Micro Signal Fusion}: Merge top-down macroeconomic signals (symbolic events) with bottom-up tick-level GNN predictions via conditional inclusion.
\end{itemize}
