\begin{tcolorbox}[
    enhanced,
    breakable,
    colback=white,
    colframe=red!70!black,
    coltitle=black,
    boxrule=1pt,
    arc=2mm,
    left=8mm,
    right=8mm,
    top=5mm,
    bottom=5mm,
    sharp corners=south,
    fonttitle=\LARGE\bfseries\sffamily,
    fontupper=\large\sffamily,
    title=\faExclamationTriangle\quad DISCLAIMER: PROCEED WITH CAUTION (AND MAYBE POPCORN)
  ]
  
  \vspace{0.5em}
  
  \textit{From Sums to Signals: A Historical Guide to Deep Learning (Featuring a Special Appearance by High-Frequency Trading)} is \textbf{NOT} peer-reviewed, NSF-funded, or even particularly well-written.

\medskip
  
  It is, however, written with the sincere hope that you’ll learn something \textit{without} wanting to gouge your eyes out from boredom.

\medskip
  
  
  \textbf{\large Let’s be clear.} I’ve done my best to be technically accurate, but I make \textbf{no guarantees} about correctness, completeness, or usefulness.  This is \textbf{edutainment}—the mythical middle ground where education and amusement try to coexist without immediately strangling each other.

\medskip
  
  \textbf{\large If you're one of the following, turn back now:}
  
  \begin{itemize}
    \item A pure academic who enjoys reading 500-page textbooks filled with Greek letters and no jokes.
    \item Someone high on the autism spectrum who craves strict formalism and zero ambiguity.
    \item A \textbf{consultant} peddling ``AI-powered disruption-as-a-service'' because I'm revealing your all your secrets.
  \end{itemize}

\medskip
  
  
  \textbf{\large Who is this for?}

\medskip
  
  Normal, curious humans—engineers, analysts, and execs with a hunch that most “AI” products are just fancy spreadsheets in trench coats.

\medskip
  
  If I can stop even one boardroom from buying a seven-figure machine learning dashboard that just runs linear regression in Comic Sans, I’ll consider this a massive win.

\medskip
  
  So relax. Laugh. Learn. Rage against the buzzword machine.
  
  \bigskip
  
  \hfill\textemdash\ \textsc{The Author}
  
  \end{tcolorbox}
  