\subsection{From Ratios to Reason: Eudoxus and the Foundations of Continuity}

The discovery of irrational numbers left Greek mathematics in an existential crisis. The Pythagoreans wanted the world to be built out of perfect whole-number ratios, but now there was this ugly, unruly thing called \( \sqrt{2} \) that refused to behave.

The Pythagoreans, in particular, would have seen the square root of 2 as a problem — a kind of anomaly or scandal in their worldview. They believed all numbers could be expressed as the ratio of whole numbers (i.e., as rational numbers). So when they discovered that \( \sqrt{2} \) couldn’t be written that way, it was deeply unsettling — not just mathematically, but philosophically and even theologically.

They would have likely referred to it as something like:

\textit{alogos} (\textgreek{ἄλογος}) — meaning ``without ratio'' or ``irrational,'' literally ``unspeakable'' or ``without reason.''

Or more generally, \textit{incommensurable} — from the Greek word \textit{asymmetros} (\textgreek{ἀσύμμετρος}), meaning two lengths that cannot be measured by the same unit (i.e., their ratio is not a rational number).

So while they didn't have the formal concept of ``irrational numbers,'' they definitely recognized the incommensurability of certain magnitudes, like the diagonal of a square to its side (which leads to \( \sqrt{2} \)).


The Greek word for \textit{incommensurable} is \textgreek{ἀσύμμετρος} (\textit{asymmetros}), 
from \textgreek{ἀ-} (``not'') and \textgreek{σύμμετρος} (``commensurate''), 
meaning ``not having a common measure.''

\textbf{Enter Eudoxus of Cnidus}: a guy who saw the whole mess and decided that maybe, just maybe, math should stop panicking and start defining things properly.

Instead of trying to force irrational numbers into a system that clearly wasn’t working, Eudoxus changed the system itself. His idea? If numbers wouldn’t play nicely, he would stop relying on numbers altogether.

\begin{figure}[H]
\centering
\begin{tikzpicture}[every node/.style={font=\footnotesize}]

% Panel 1 — Pythagorean panicking
\comicpanel{0}{4}
  {Pythagorean}
  {Eudoxus}
  {\textbf{Pythagorean:} \(\sqrt{2}\) won’t reduce to a ratio!  
We’ve tried every fraction!}
  {(0,-0.5)}

% Panel 2 — Eudoxus suggesting abstraction
\comicpanel{6.5}{4}
  {Pythagorean}
  {Eudoxus}
  {\textbf{Eudoxus:} Have you tried comparing it to...  
an idea instead?}
  {(0,-0.5)}

% Panel 3 — Pythagorean confused
\comicpanel{0}{0}
  {Pythagorean}
  {Eudoxus}
  {\textbf{Pythagorean:} I’m sorry—what?  
What does that even mean?}
  {(0,0.8)}

% Panel 4 — Eudoxus chill
\comicpanel{6.5}{0}
  {Pythagorean}
  {Eudoxus}
  {\textbf{Eudoxus:} Math isn’t broken.  
Your expectations are.}
  {(0,0.8)}

\end{tikzpicture}
\caption{When numbers got messy, Eudoxus invented abstraction. The rest of math is still catching up.}
\end{figure}