\subsection{Information Geometry: A Statistical Shadow of Kepler's Law}

What if Kepler’s Second Law—the conservation of swept area—had an analogue beyond celestial mechanics?  
What if, instead of tracing planets around the Sun, we traced trajectories through a space of \textit{probability distributions}?

Enter \textbf{information geometry}.

In information geometry, we treat families of probability distributions as points on a manifold.  
Each parameterized model—each choice of mean, variance, or other statistical parameter—defines a coordinate on this manifold.

But this space is not flat.  
It comes equipped with a natural curvature, measured by the \textbf{Fisher information metric}:

\[
g_{ij}(\theta) = \mathbb{E}\left[ \frac{\partial \log p(x \mid \theta)}{\partial \theta^i} \frac{\partial \log p(x \mid \theta)}{\partial \theta^j} \right]
\]

This metric turns statistical inference into a geometric journey:  
Every learning step is a move across this manifold; every update traces a curve through parameter space.

\medskip

\noindent
\textbf{And here’s the key:}  
Just as Kepler’s Second Law conserves area in phase space,  
certain flows in information geometry conserve an “area” in statistical space—  
a volume element determined by the Fisher information.

\[
dV = \sqrt{\det g(\theta)} \, d\theta^1 \wedge \dots \wedge d\theta^n
\]

Under natural gradient flows—introduced by Shun-ichi Amari—this information volume behaves analogously to symplectic area in mechanics.  
The learning trajectory sweeps out “equal information” in “equal effort” steps.

\begin{quote}
\textbf{The orbit of a learner mirrors the orbit of a planet.  
Each preserves a hidden measure:  
planets conserve geometric area; learners conserve informational divergence.}
\end{quote}

\begin{tcolorbox}[colback=blue!5!white, colframe=blue!75!black, title={Analogy Table}]
\begin{tabular}{ll}
\textbf{Celestial Mechanics} & \textbf{Information Geometry} \\
\hline
Phase space coordinates $(q,p)$ & Parameter space coordinates $\theta$ \\
Symplectic 2-form $\omega$ & Fisher information metric $g_{ij}$ \\
Swept area $\int_\gamma \omega$ & Information volume $\int_\gamma \sqrt{\det g}\, d\theta$ \\
Conserved angular momentum & Conserved information gain \\
Kepler’s Second Law & Natural gradient flow invariance \\
\end{tabular}
\end{tcolorbox}

In this analogy, an inference algorithm—like stochastic gradient descent or Bayesian updating—does not just move arbitrarily.  
It moves along paths constrained by the underlying information geometry,  
preserving informational invariants the way a planet preserves angular momentum.

\begin{quote}
\textit{Learning is not random wandering—it is a geodesic through belief space,  
respecting the curvature induced by statistical structure.}
\end{quote}

\subsubsection*{A Unified View}

We can now glimpse a deeper unity:

\begin{itemize}
    \item In mechanics, dynamics preserve symplectic area: $d\omega = 0$.
    \item In information geometry, learning preserves informational volume: $d(\sqrt{\det g}) = 0$ along natural flows.
\end{itemize}

Both reflect an invariant 2-form—whether geometric or statistical.

Both suggest that motion, whether physical or inferential,  
is shaped not by arbitrary rules, but by the hidden geometry of the space itself.

\begin{tcolorbox}[colback=gray!5!white, colframe=gray!50!black, title={Summary}]
Kepler’s Law speaks not only to celestial orbits,  
but to a broader principle:

\begin{quote}
\textbf{Wherever systems evolve, there exists a geometry that constrains them,  
and an invariant that measures what they cannot lose.}
\end{quote}
\end{tcolorbox}

In this light, Kepler’s discovery is not merely an astronomical law.  
It is an early glimpse of a universal theme:

\begin{quote}
\textit{Conservation is geometry’s fingerprint on evolution—  
whether in the heavens, or in the space of belief.}
\end{quote}
