\subsection{What Happens When a Theologian Does Too Much Math?}

If Galileo’s crime was mocking the Pope, Johannes Kepler’s was simply existing at the wrong time. A devout Lutheran, an obsessive mathematician, and a man who thought the universe was built on divine geometry, Kepler \textbf{just wanted to do horoscopes and prove that God was a geometer.} Instead, he destroyed Aristotle’s final hope of keeping planetary motion simple and got himself tangled in religious wars, witch trials, and his own mentor’s murder.

Before Kepler, everyone was still clinging to perfect circles. The idea that planets must move in circles was deeply ingrained in Aristotelian physics and medieval theology. Why? Because circles are perfect. And God wouldn’t make something as messy as an ellipse.

Kepler, however, wasn’t interested in what people wanted planetary motion to look like. He had actual data.

Kepler was hired by \textbf{Tycho Brahe}, an astronomer whose personality could be described as \textit{"drunk Viking who traded Thor for Jesus."} Brahe had a nose made of metal (lost in a duel), a pet moose (which died from drinking too much beer), lived a life of sexual immorality (essentially never officially married his wife because his high social status made it illegal to marry a commoner),  and had \textbf{the most accurate astronomical data in history} (which he refused to share).

When Brahe died mysteriously in 1601 (possibly poisoned, but let’s not point fingers), Kepler inherited his data. And with years of obsessive calculations, he formulated three laws that finally shattered Aristotelian physics.

But it was \textbf{Kepler’s Second Law} that changed everything.

\begin{figure}[H]
\centering

% === First row ===
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Tycho's Father}
    {}
    {\footnotesize You can’t marry her. She’s a commoner. It’s an affront to the family name.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Nobility: allergic to love.}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.45\textwidth}
  \centering
  \begin{tikzpicture}
    \comicpanel{0}{0}
      {Minister}
      {}
      {\footnotesize Wed to a commoner? That’s being unequally yoked, in every sense.}
      {(0,-0.6)}
  \end{tikzpicture}
  \caption*{Sanctity denied by class and scripture.}
  \end{subfigure}
  

\vspace{1em}

% === Second row ===
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Family Banker}
    {}
    {\footnotesize That marriage jeopardizes your inheritance. It’s an affront to generational wealth.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Romance vs. compounding interest.}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Tycho Brahe}
    {}
    {\footnotesize Hey babe, you can be my live-in girlfriend. Forever.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Living in sin... with style.}
\end{subfigure}

\caption{Tycho Brahe: Losing his nose in a duel was somehow not the most dramatic part of his life.}
\end{figure}

