\subsection{The Controversy: Kepler vs. The Universe}

Why was this a problem for Aristotle?

Because Aristotle’s physics said motion required a constant force. If planets were moving faster and slower without anyone pushing them, something was terribly wrong. \textbf{Kepler had just killed another fundamental assumption of Aristotelian motion}.

Which brings us to the real controversy.

Kepler didn’t just break the rules of celestial motion: he broke the rules of polite intellectual society.

\begin{itemize} 
	\item \textbf{The Catholics hated him} because he was Protestant. 
	\item \textbf{The Protestants hated him} because he wasn’t their \textit{kind} of Protestant. 
	\item \textbf{The Aristotelians hated him} because he said their perfect heavens were lurching around like drunkards. 
\end{itemize}

And just when things couldn’t get worse, his personal life problems were compounded by misfortune: \textbf{his mother was put on trial for witchcraft}. (Because obviously, if your son can predict planetary motion, your whole family must be in league with Satan.)

Kepler barely saved her from being burned at the stake, but he couldn’t save his own career. His insights, like the idea that you could use area to describe motion, were simply too strange, too early, and too threatening to the world’s tidy beliefs.

Kepler’s use of areas to describe motion was a game-changer. It wasn’t just about astronomy: it was about \textbf{introducing the idea that motion and change could be calculated dynamically}.

Of course, Kepler didn’t get to see himself become a legend. He died broke, excommunicated, and largely ignored. 




\begin{figure}[H]
\centering

% === First row ===
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Catholic Priest}
    {Kepler}
    {\footnotesize Heresy! You read the Bible... without a priest!}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Theological autonomy: forbidden.}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Lutheran Pastor}
    {Kepler}
    {\footnotesize Heresy! You read the Bible... wrong.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Interpretation: not a solo sport.}
\end{subfigure}

\vspace{1em}

% === Second row ===
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Cosmologist}
    {Kepler}
    {\footnotesize Heresy! You tried to fix the heavens with math, and not meaning.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Orthodoxy dies in elliptical orbits.}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Tycho Brahe}
    {Kepler}
    {\footnotesize I once got in a duel over math and lost my nose. You’ll be fine.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Mentorship: Viking edition.}
\end{subfigure}

\caption{Kepler’s crime: using math to describe the universe. Everyone else’s crime: being alive in the 1600s.}
\end{figure}

