\subsection{Newton’s Fluxions: From Geometry to Motion (1670s)}

  Isaac Newton didn’t just invent calculus: he invented a new way to think about time itself.
  
  In the \textit{Principia}, he draws a sharp distinction between two kinds of time:

  \begin{quote}
  Absolute, true, and mathematical time, of itself, and from its own nature, flows equably without relation to anything external...
  \end{quote}
  
  For Newton, time wasn’t just a coordinate. It was a cosmic river—unseen, unchanging, and universal. All motion, all change, all flux happened within this background flow. It didn’t depend on clocks or planets or observers. Time just \textit{was}, always ticking along in the background like some divine metronome.
  
  This view shaped his entire mathematical framework. The derivative—what he called the \textbf{fluxion}—wasn’t just a slope on a graph. It was a rate of change \textit{with respect to absolute time}. Every evolving quantity had a speed. Every curve had a motion. And every motion happened within the silent rhythm of Newton’s eternal clock.

  Where others saw geometry, Newton saw \textbf{kinematics}. Where others measured lengths, he measured \textbf{change}. And at the center of it all was time—not as an illusion, but as the most real thing in the universe.

    Before Newton, motion was geometry's unruly cousin — more art than science, more epicycle than equation. Astronomers charted paths. Mathematicians constructed curves. But few asked the deeper question: \textit{how does motion unfold through time?} \textbf{Kinematics} changed that.

  \begin{tcolorbox}[colback=gray!5!white, colframe=black, title=\textbf{Historical Sidebar: Newton and Augustine — Time Between Eternity and Motion}, fonttitle=\bfseries, arc=1.5mm, boxrule=0.4pt]

    When Isaac Newton described \textit{absolute time} as flowing “of itself, from its own nature, equably without relation to anything external,” he wasn’t just making a mathematical statement: he was echoing a much older philosophical tradition.

    \medskip
    
    Over a thousand years earlier, \textbf{St. Augustine} grappled with the mystery of time in his \textit{Confessions}. Augustine famously asked:
    
    \begin{quote} \textit{“What then is time? If no one asks me, I know; if I wish to explain it to one that asketh, I know not.”} \end{quote}
    
    For Augustine, time was inseparable from the soul’s perception. He argued that past and future exist only as present memory and anchored in human consciousness. God, by contrast, exists in a timeless eternity where all moments are present at once.

    \medskip
    
    While Augustine located time within the mind’s experience, Newton externalized it. Time, for Newton, was not psychological: it was a universal \textbf{divine framework} set by God Himself. In fact, Newton’s absolute time can be seen as a mathematical counterpart to Augustine’s theological eternity: both are immutable, unaffected by the material world.

    \medskip
    
    But where Augustine saw time as a measure of the soul’s journey through creation, Newton transformed it into a precise, ticking backdrop against which \textit{motion} could be quantified.

    \medskip
    
    \begin{center}
      \begin{tabular}{c c c}
      \textbf{Augustine} &  & \textbf{Newton} \\
      \hline
      Time as inner experience & $\longleftrightarrow$ & Time as cosmic constant \\
      Bound to consciousness   &                      & Independent of observers \\
      Eternity as divine present &                    & Absolute time as divine clockwork \\
      \end{tabular}
      \end{center}
    
    \medskip

    Yet both men agreed on one thing: \textbf{time was real, and it flowed according to divine principles}. For Augustine, those principles were psychological; for Newton, they were mechanical, but still underpinned by creator God.
    
    \end{tcolorbox}
  

    \medskip
    
    In the 17th century, a new discipline emerged — one that stripped away causes and forces and focused only on motion itself. It didn’t ask \textit{why} things moved, only \textit{how fast}, \textit{in what direction}, and \textit{how that changed over time}.

    \textbf{Key Developments:}
    

    \begin{itemize}
      \item \textbf{Galileo Galilei} studied uniformly accelerated motion and laid the groundwork for describing motion using time, distance, and velocity — even before calculus.
      \item \textbf{Isaac Newton} took this further, developing calculus as a tool for describing how quantities change — giving us acceleration, jerk, and higher derivatives as measurable realities.
      \item Kinematics became the scaffolding for Newton’s deeper theory: \textbf{dynamics}, which added force and mass into the mix.
    \end{itemize}
    
    Kinematics was the first major shift in physics to treat time as a variable rather than a stage. It formalized the idea that motion is not just spatial, but \textbf{temporal}. Every modern theory of motion — from relativity to quantum mechanics — builds on this insight.
    
    \begin{quote}
    Kinematics didn’t explain the universe.  It gave us the vocabulary to describe its dance.
    \end{quote}

To Newton, quantities were not static magnitudes, but flowing values—changing continuously with time. Instead of drawing triangles on curves, he imagined how a point moved along a curve, and how its position evolved. This shift—from shape to speed, from geometry to kinematics—transformed the mathematics of the curve into the mathematics of change itself.

He denoted this with a dot over the variable:

\[
\dot{x} = \frac{o}{\bar{o}}
\]

where:

\begin{itemize}
    \item \( o \) is an \textbf{infinitely small increment} of time.  
    \item \( \bar{o} \) is the corresponding change in \( x \).
\end{itemize}


  

Unlike his predecessors, Newton did not merely approximate motion: he embedded it into the foundations of his system. Tangents were consequences of how quantities flowed.

  Newton didn’t invent acceleration—that was Galileo, who noticed that falling objects don’t just move; they \textit{accelerate}. But Galileo stopped at observation. Newton gave it a cause—and a calculus.
  
  \textbf{Step one: Define acceleration in terms of fluxions.}  
  Using his own notation, Newton described acceleration as the \textit{fluxion of velocity}, written as:
  \[
  \dot{v}
  \]
  
  \textbf{Step two: Define momentum.}  

  He understood that the quantity of motion—what we now call momentum—was proportional to mass and velocity:
  \[
  p = m v
  \]
  
  \textbf{Step three: Relate force to the change of momentum.}  

  In the \textit{Principia}, Newton writes:

  \begin{quote}
  \textit{“The change of motion is proportional to the motive force impressed.”}
  \end{quote}

  Here, “motion” means momentum. So mathematically:

  \[
  F = \dot{p}
  \]
  
  Now, assuming mass is constant (which Newton generally did), this becomes:

  \[
  F = \dot{(mv)} = m\dot{v}
  \]
  
  \textbf{And there it is.} Not \( F = ma \), but \( F = m\dot{v} \): force as the fluxion of momentum.
  
  \textbf{Galileo saw that velocity changed with time. Newton gave that change a cause—and a law.}
  
  By linking force to motion through fluxions, Newton wasn’t just describing the world. He was explaining it—on Earth, in the heavens, and everywhere time flows.


\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=2]
    % Draw the curve
    \draw[thick, domain=-1:1.2, smooth, variable=\x] plot ({\x},{\x*\x});

    % Draw infinitesimal motion
    \draw[->, thick] (0.5, 0.25) -- (0.7, 0.49) node[midway, above] {\small $o$};
    \draw[thick] (0.5, 0.25) -- (0.7, 0.25);
    \draw[thick] (0.7, 0.25) -- (0.7, 0.49);
    \node[below] at (0.6,0.25) {\small $\bar{o}$};
    
    % Draw tangent line
    \draw[dashed] (0.2, 0.04) -- (1.1, 0.91) node[above] {\small Tangent Line};

\end{tikzpicture}

\vspace{0.5em}
\caption{\small Newton's conception of calculus was rooted in geometry and motion. He imagined a point moving along a curve, where the instantaneous rate of change — the fluxion — could be represented as the ratio between an infinitesimal vertical change ($o$) and an infinitesimal horizontal change ($\bar{o}$). The resulting quantity corresponds to the slope of the tangent line at that point. This approach reflects Newton’s view of change as inherently tied to physical movement through time.}
\end{figure}



