\section{Leibniz: Calculus by Rules, Not Just Results (1675–1700)}

While Newton built calculus through series and geometry, \textbf{Gottfried Wilhelm Leibniz} turned it into a formal language.  
He didn’t just want to solve problems—he wanted to mechanize the solving itself.

Leibniz introduced the notation we still use today: \( \frac{dy}{dx} \), \( \int \), and the differential \( dx \).  
But more than that, he developed a systematic method for computing derivatives by applying symbolic rules.

\subsection{Differentiation Becomes a Calculus}

Where Newton worked through fluents and fluxions tied to time, Leibniz treated variables algebraically and operations procedurally.  
He discovered—and codified—general rules for how to differentiate powers, products, quotients, and even early transcendental functions like logarithms and exponentials:

\[
\frac{d}{dx} x^n = n x^{n-1} \qquad \text{(for any positive integer \( n \))}
\]

\[
\frac{d}{dx} \log x = \frac{1}{x}, \qquad \frac{d}{dx} (e^x) = e^x
\]

\[
\frac{d}{dx}(uv) = u'v + uv' \qquad \text{(Product Rule)}
\quad\quad
\frac{d}{dx} \left( \frac{u}{v} \right) = \frac{u'v - uv'}{v^2} \qquad \text{(Quotient Rule)}
\]

Leibniz’s true genius was in how he treated these operations as \textit{syntactic}—a grammar of change.  
You didn’t need to visualize motion or sum areas. You could manipulate symbols according to rules, like playing algebra in fast-forward.

\subsection{From Curves to Calculators}

This perspective was radical. It laid the foundation for treating calculus like a machine—an idea that would echo through later thinkers like Laplace, Lagrange, and eventually into the logic of modern computers.

\vspace{1em}

\begin{center}
\textit{If Newton saw calculus in the curves of nature, Leibniz heard it in the grammar of reason.}
\end{center}


\subsection{A New Kind of Equation: Relations of Change}

Leibniz didn’t just invent notation—he opened the door to a new kind of equation.

By symbolizing derivatives as \( \frac{dy}{dx} \), he made it possible to write \textbf{equations where the unknown wasn’t a number, but a function whose rate of change followed certain rules}. These became known as \textbf{differential equations}: equations that relate a function to its derivatives.

For Leibniz, differential equations were a natural extension of calculus. If differentiation measured change, and integration summed change, then a differential equation was a statement about how a quantity \emph{must change}—a recipe for motion, growth, or interaction.

This wasn’t just symbolic cleverness. It created a new mathematical language for describing the world. Forces, orbits, velocities, flows—anything changing over time or space could now be encoded as an equation involving derivatives.

Leibniz’s framework laid the foundation for later generations of mathematicians and physicists, who would turn these symbolic expressions into tools for solving real-world problems.

\vspace{1em}

\begin{center}
\textit{Leibniz gave calculus its grammar—and with it, a way to write the laws of change themselves.}
\end{center}